---
title: "End-to-end WCC + peak-pick speedup: cumc, cumcuda vs legacy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end WCC + peak-pick speedup: cumc, cumcuda vs legacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  cache    = FALSE
)
```

This vignette measures the end-to-end speedup of the recent `wcc`
rewrite — **including the time to compute the WCC grids themselves** —
on a single simulated dataset chosen to be favorable to the GPU
backend (many dyads, moderately long series). The comparison is:

- **Legacy pipeline** — original C backend for the WCC
  (`wccCalc(..., method = "c")`, i.e. `windcross`), followed by the
  original per-row `loess() + spline()` peak picker, embedded verbatim
  below as `wccPeakPickLegacy`. Both steps are looped over dyads in R.
- **`cumc` pipeline** — new cumulative-sum C backend, batched
  (`wccCalcBatch(..., method = "cumc")`), followed by the new
  matrix-smoother peak picker (`wccPeakPickBatch(..., method = "cumc")`).
- **`cumcuda` pipeline** — new cumulative-sum CUDA backend, batched
  (`wccCalcBatch(..., method = "cumcuda")`), followed by the CUDA
  batched peak picker (`wccPeakPickBatch(..., method = "cumcuda")`).
- **`cumcuda` FP32 pipeline** — same as above with
  `precision = "single"`: all device arithmetic runs in FP32, which on
  consumer GPUs (whose FP64 throughput is 1/32–1/64 of FP32) unlocks
  most of the card. Results are accurate to roughly `1e-5` instead of
  machine precision.
- **fused pipeline** — `wccCalcPeakBatch(..., method = "cumcuda")`:
  both stages in one call, with the WCC grids resident on the GPU
  throughout. The two-step GPU pipeline downloads the full
  `nRow × nCol × P` grid array to R, zeroes the NAs there, and uploads
  it again for the smoother GEMM; at this problem size that round trip
  dominates the runtime, which is also why FP32 alone barely moves the
  two-step numbers. The fused call transfers only the small
  `nRow × P` index and value matrices. Timed in FP64 and FP32.

Both stages of each pipeline are inside the `system.time()` block, so
the reported numbers reflect what a user of `wccFindDyadParam` or a
similar batched analysis will actually feel.

```{r libraries}
library(wcc)
set.seed(20260701)
```

## The legacy reference

To make the accuracy claim verifiable inside the vignette, we embed the
original per-row peak-pick implementation as `wccPeakPickLegacy`. It is
bit-identical to the `wccPeakPick` that shipped in `wcc` prior to the
refactor, minus the dead plotting code. The legacy pipeline then loops
over dyads calling `wccCalc(..., method = "c")` and passing the result
to `wccPeakPickLegacy`.

```{r legacy_defs}
wccPeakPickLegacy <- function(tAllCor = NA, Lsize = 8, pspan = .25, type = "Max") {
    colLen <- length(tAllCor[1, ])
    rowLen <- length(tAllCor[, 1])
    mx <- rep(NA, (2 * colLen - 1))
    tIndex <- rep(NA, rowLen)
    tValue <- rep(NA, rowLen)
    findMin <- (type == "Min" || type == "min")
    for (rowNo in seq_len(rowLen)) {
        if (any(is.na(tAllCor[rowNo, ]))) next
        tCor <- tAllCor[rowNo, ]
        t1 <- predict(loess(tCor ~ c(1:colLen), degree = 2, span = pspan))
        t2 <- spline(c(1:colLen), t1, n = (2 * colLen - 1))$y
        windowWidth <- 0
        lookAhead <- 0
        for (j in 1:(colLen - 1)) {
            windowWidth <- windowWidth + 1
            tSelect <- (colLen - windowWidth):(colLen + windowWidth)
            mx[j] <- if (findMin) min(t2[tSelect], na.rm = TRUE) else max(t2[tSelect], na.rm = TRUE)
            if (j == 1) mmx <- mx[j]
            else {
                better <- if (findMin) (mx[j] < mmx) else (mx[j] > mmx)
                if (better) { lookAhead <- 0; mmx <- mx[j] }
                else {
                    lookAhead <- lookAhead + 1
                    if (lookAhead >= Lsize) break
                }
            }
        }
        Index <- match(mmx, t2[tSelect]) + tSelect[1] - 1
        position <- Index - colLen
        if (position > (colLen - Lsize - 1) || position < (-(colLen - Lsize - 1))) {
            tIndex[rowNo] <- NA
            tValue[rowNo] <- NA
        } else {
            tIndex[rowNo] <- position
            tValue[rowNo] <- mmx
        }
    }
    if (findMin) list(minIndex = tIndex, minValue = tValue)
    else list(maxIndex = tIndex, maxValue = tValue)
}

legacyPipeline <- function(arr1, arr2, wMax, tMax, Lsize, pspan) {
    D <- nrow(arr1)
    index <- vector("list", D)
    value <- vector("list", D)
    for (p in seq_len(D)) {
        grid <- wccCalc(arr1[p, ], arr2[p, ], wMax = wMax, tMax = tMax,
                        method = "c")
        grid[is.na(grid)] <- 0
        pp <- wccPeakPickLegacy(grid, Lsize = Lsize, pspan = pspan, type = "Max")
        index[[p]] <- pp$maxIndex
        value[[p]] <- pp$maxValue
    }
    list(index = do.call(cbind, index),
         value = do.call(cbind, value))
}
```

## A dataset favorable to CUDA

CUDA's batched backend pays off when there are many dyads (large
`P = D`) and moderately long series. The GPU launches one thread block
per (pair, output column) for WCC, and one strided-batched GEMM plus
one thread per (slab, row) for peak picking. Both stages saturate the
GPU only above a minimum problem size.

```{r params}
D      <- 40      # number of dyads (= number of slabs P)
n      <- 2000    # samples per series
wMax   <- 60      # window size
tMax   <- 60      # maximum lag
Lsize  <- 8
pspan  <- 0.25
```

Simulate `D` pairs of noisy sinusoids and stack them as `D x n`
matrices.

```{r simulate}
makeSeries <- function(n) {
    sin(seq_len(n) / runif(1, min = 5, max = 20)) + rnorm(n, sd = 0.5)
}

arr1 <- t(sapply(seq_len(D), function(i) makeSeries(n)))
arr2 <- t(sapply(seq_len(D), function(i) makeSeries(n)))
```

## Detect the CUDA build

```{r cuda_probe}
cudaAvailable <- tryCatch({
    wccCalc(rnorm(200), rnorm(200), wMax = 20, tMax = 20, method = "cumcuda")
    TRUE
}, error = function(e) FALSE)
cat("CUDA backend available:", cudaAvailable, "\n")
```

## Warm-up

Every backend is called once on a tiny input before measurement so the
matrix-smoother cache (`wccSmoothMatrix`), any lazy compilation, and
CUDA context setup do not contaminate timings.

```{r warm}
invisible(legacyPipeline(arr1[1:2, , drop = FALSE], arr2[1:2, , drop = FALSE],
                          wMax = wMax, tMax = tMax, Lsize = Lsize, pspan = pspan))
invisible({
    g <- wccCalcBatch(arr1[1:2, ], arr2[1:2, ], wMax = wMax, tMax = tMax,
                      method = "cumc")
    g[is.na(g)] <- 0
    wccPeakPickBatch(g, Lsize = Lsize, pspan = pspan, type = "Max",
                     method = "cumc")
})
if (cudaAvailable) {
    invisible({
        g <- wccCalcBatch(arr1[1:2, ], arr2[1:2, ], wMax = wMax, tMax = tMax,
                          method = "cumcuda")
        g[is.na(g)] <- 0
        wccPeakPickBatch(g, Lsize = Lsize, pspan = pspan, type = "Max",
                         method = "cumcuda")
    })
    invisible({
        g <- wccCalcBatch(arr1[1:2, ], arr2[1:2, ], wMax = wMax, tMax = tMax,
                          method = "cumcuda", precision = "single")
        g[is.na(g)] <- 0
        wccPeakPickBatch(g, Lsize = Lsize, pspan = pspan, type = "Max",
                         method = "cumcuda", precision = "single")
    })
    invisible(wccCalcPeakBatch(arr1[1:2, ], arr2[1:2, ], wMax = wMax, tMax = tMax,
                               Lsize = Lsize, pspan = pspan, type = "Max",
                               method = "cumcuda"))
    invisible(wccCalcPeakBatch(arr1[1:2, ], arr2[1:2, ], wMax = wMax, tMax = tMax,
                               Lsize = Lsize, pspan = pspan, type = "Max",
                               method = "cumcuda", precision = "single"))
}
```

## Timing the three pipelines end-to-end

Each timing block covers **both** the WCC computation and the peak
picking — the two stages a user runs to get from raw series to per-
dyad `(lag, correlation)` peaks.

**Legacy: `wccCalc(method = "c")` + per-row `loess + spline`, looped**

```{r time_legacy}
tLegacy <- system.time({
    legacyOut <- legacyPipeline(arr1, arr2,
                                 wMax = wMax, tMax = tMax,
                                 Lsize = Lsize, pspan = pspan)
})["elapsed"]
tLegacy
```

**`cumc`: `wccCalcBatch(method = "cumc")` + `wccPeakPickBatch(method = "cumc")`**

```{r time_cumc}
tCumc <- system.time({
    gCumc <- wccCalcBatch(arr1, arr2, wMax = wMax, tMax = tMax, method = "cumc")
    gCumc[is.na(gCumc)] <- 0
    outCumc <- wccPeakPickBatch(gCumc, Lsize = Lsize, pspan = pspan,
                                 type = "Max", method = "cumc")
})["elapsed"]
tCumc
```

**`cumcuda`: `wccCalcBatch(method = "cumcuda")` + `wccPeakPickBatch(method = "cumcuda")`**

```{r time_cumcuda}
if (cudaAvailable) {
    tCumcuda <- system.time({
        gCumcuda <- wccCalcBatch(arr1, arr2, wMax = wMax, tMax = tMax,
                                  method = "cumcuda")
        gCumcuda[is.na(gCumcuda)] <- 0
        outCumcuda <- wccPeakPickBatch(gCumcuda, Lsize = Lsize, pspan = pspan,
                                        type = "Max", method = "cumcuda")
    })["elapsed"]
} else {
    tCumcuda <- NA_real_
    outCumcuda <- NULL
}
tCumcuda
```

**`cumcuda` FP32: the same GPU pipeline with `precision = "single"`**

```{r time_cumcuda32}
if (cudaAvailable) {
    tCumcuda32 <- system.time({
        gCumcuda32 <- wccCalcBatch(arr1, arr2, wMax = wMax, tMax = tMax,
                                    method = "cumcuda", precision = "single")
        gCumcuda32[is.na(gCumcuda32)] <- 0
        outCumcuda32 <- wccPeakPickBatch(gCumcuda32, Lsize = Lsize, pspan = pspan,
                                          type = "Max", method = "cumcuda",
                                          precision = "single")
    })["elapsed"]
} else {
    tCumcuda32 <- NA_real_
    outCumcuda32 <- NULL
}
tCumcuda32
```

**Fused: `wccCalcPeakBatch(method = "cumcuda")` — grid stays on the GPU**

```{r time_fused}
if (cudaAvailable) {
    tFused <- system.time({
        outFused <- wccCalcPeakBatch(arr1, arr2, wMax = wMax, tMax = tMax,
                                     Lsize = Lsize, pspan = pspan,
                                     type = "Max", method = "cumcuda")
    })["elapsed"]
} else {
    tFused <- NA_real_
    outFused <- NULL
}
tFused
```

**Fused FP32: `wccCalcPeakBatch(..., precision = "single")`**

```{r time_fused32}
if (cudaAvailable) {
    tFused32 <- system.time({
        outFused32 <- wccCalcPeakBatch(arr1, arr2, wMax = wMax, tMax = tMax,
                                       Lsize = Lsize, pspan = pspan,
                                       type = "Max", method = "cumcuda",
                                       precision = "single")
    })["elapsed"]
} else {
    tFused32 <- NA_real_
    outFused32 <- NULL
}
tFused32
```

## Correctness

The two new pipelines should reproduce the legacy output up to
floating-point noise. Peak indices are integer-valued and must match
exactly; peak values sit at machine epsilon on the CPU path and at
low-order double precision on the GPU path (the strided-batched GEMM
accumulates in a different order than the R + BLAS smoother).

```{r verify}
maxAbsDiff <- function(a, b) {
    stopifnot(length(a) == length(b))
    naA <- is.na(a); naB <- is.na(b)
    stopifnot(all(naA == naB))
    max(abs(a - b), na.rm = TRUE)
}

cat(sprintf("legacy vs cumc:    max |Δ index| = %.2e   max |Δ value| = %.2e\n",
            maxAbsDiff(legacyOut$index, outCumc$index),
            maxAbsDiff(legacyOut$value, outCumc$value)))

if (cudaAvailable) {
    cat(sprintf("legacy vs cumcuda: max |Δ index| = %.2e   max |Δ value| = %.2e\n",
                maxAbsDiff(legacyOut$index, outCumcuda$index),
                maxAbsDiff(legacyOut$value, outCumcuda$value)))
}
```

The FP32 pipeline trades machine precision for speed: peak *values*
agree with the FP64 result to about `1e-5`, and peak *indices* match
except on near-ties, where the picked position can shift by one
half-step.

```{r verify32}
if (cudaAvailable) {
    idxShift <- sum(abs(outCumcuda32$index - legacyOut$index) > 0, na.rm = TRUE)
    cat(sprintf("legacy vs cumcuda FP32: max |Δ value| = %.2e   shifted indices: %d/%d\n",
                max(abs(legacyOut$value - outCumcuda32$value), na.rm = TRUE),
                idxShift, sum(!is.na(legacyOut$index))))
}
```

The fused pipeline computes the same quantities as the two-step GPU
pipeline (with the `NA → 0` substitution applied on the device), so it
must agree with it — exactly in FP64, and up to the same near-tie
caveat in FP32.

```{r verify_fused}
if (cudaAvailable) {
    cat(sprintf("legacy vs fused:   max |Δ index| = %.2e   max |Δ value| = %.2e\n",
                maxAbsDiff(legacyOut$index, outFused$index),
                maxAbsDiff(legacyOut$value, outFused$value)))
    idxShiftF <- sum(abs(outFused32$index - legacyOut$index) > 0, na.rm = TRUE)
    cat(sprintf("legacy vs fused FP32: max |Δ value| = %.2e   shifted indices: %d/%d\n",
                max(abs(legacyOut$value - outFused32$value), na.rm = TRUE),
                idxShiftF, sum(!is.na(legacyOut$index))))
}
```

## Speedup summary

```{r summary_table, results='asis'}
tab <- data.frame(
    pipeline = c("legacy: windcross + loess+spline peak-pick (looped)",
                 "cumc: wccCalcBatch + wccPeakPickBatch (CPU)",
                 "cumcuda: wccCalcBatch + wccPeakPickBatch (GPU, FP64)",
                 "cumcuda FP32: same with precision = \"single\"",
                 "fused: wccCalcPeakBatch (GPU, FP64, grid stays on device)",
                 "fused FP32: wccCalcPeakBatch with precision = \"single\""),
    elapsed_s = c(tLegacy, tCumc, tCumcuda, tCumcuda32, tFused, tFused32),
    speedup_vs_legacy = c(1, tLegacy / tCumc,
                          if (cudaAvailable) tLegacy / tCumcuda else NA_real_,
                          if (cudaAvailable) tLegacy / tCumcuda32 else NA_real_,
                          if (cudaAvailable) tLegacy / tFused else NA_real_,
                          if (cudaAvailable) tLegacy / tFused32 else NA_real_)
)
knitr::kable(tab, digits = c(0, 2, 1),
             caption = sprintf(
                 "End-to-end WCC + peak-pick time on D=%d dyads, n=%d, wMax=tMax=%d",
                 D, n, wMax))
```

Two sources of speedup are stacked in these numbers:

1. **`wccCalc` → `wccCalcBatch(method = "cumc")`**. The cumulative-sum
   algorithm makes each WCC cell an `O(1)` prefix-sum difference, so
   the total cost per dyad is independent of `wMax`. On top of that,
   the batched form computes each series' prefix sums once and shares
   them across all pairs, and parallelizes pair iteration with OpenMP.
2. **legacy peak-pick → `wccPeakPickBatch(method = "cumc")`**. The
   `loess(degree = 2) + spline` smoother at fixed predictor positions
   is a fixed linear operator: a single `(2·colLen − 1) × colLen`
   matrix that depends only on `(colLen, pspan)`. `wccSmoothMatrix`
   builds and caches it; all rows of all slabs are then smoothed with
   one matrix product instead of one `loess` fit per row.

`cumcuda` adds a third layer: both stages run on the GPU, with
per-series prefix-sum scans in shared memory, one CUDA thread block
per (pair, output column) for the WCC kernel, one
`cublasDgemmStridedBatched` call for the batched smoother, and one
thread per (slab, row) for the peak search. Only the small
`P × nRow` index and value arrays cross back to the host.

`precision = "single"` adds a fourth layer on top of `cumcuda`: the
inputs are narrowed to FP32 on the host (halving PCIe traffic and
device memory), all kernels run in float, and the smoother GEMM becomes
`cublasSgemmStridedBatched`. On consumer cards, where FP64 units are
scarce, this raises the arithmetic ceiling — but as the two-step
numbers above show, arithmetic is not the bottleneck at this problem
size, so FP32 alone changes little.

The **fused** `wccCalcPeakBatch` removes that bottleneck. In the
two-step GPU pipeline the full grid array (`nRow × nCol × P` doubles,
tens of MB here) is downloaded to R, scanned for NAs, and uploaded
again — three passes over PCIe plus two host-side loops, for data the
GPU already had. The fused call keeps the grids resident on the device
(zero-variance cells are written as `0` directly, matching the
`g[is.na(g)] <- 0` convention), runs the smoother GEMM and peak search
in place, and downloads only the `nRow × P` index and value matrices.
The backend also reuses its device buffers, cuBLAS handle, and pinned
transfer staging across calls, which matters for repeated invocations
in a parameter search. With transfers out of the way, FP32 finally has
something to accelerate, which is why the fused FP32 row shows the
largest gain.

## What determines your numbers

Absolute timings depend heavily on:

- **`D` (number of dyads)** — the CUDA backend needs enough parallel
  work to saturate the device. Below `D ≈ 10` the CPU batched path
  may already be competitive; above `D ≈ 100` the GPU tends to pull
  ahead.
- **`n` (series length)** — longer series make the CUDA per-series
  prefix-sum scan more efficient relative to launch overhead.
- **CPU threads for `cumc`** — the batched C backend uses OpenMP over
  the pair loop. On a single-core CI runner the CPU speedup is
  smaller but the GPU speedup is roughly unchanged.
- **`wMax ≤ 2048`** — the CUDA kernel tiles the time axis in
  shared-memory blocks of 2048 samples; larger windows fall back to a
  host error. This is not a limit on the CPU paths.

Rule of thumb for a parameter-search workload
(`wccFindDyadParam` with many surrogates):

- ≤ 5 dyads, single parameter: `method = "c"` is fine.
- 20–100 dyads, moderate parameter grid: `method = "cumc"` is the
  right default.
- 100+ dyads, dense parameter grid, CUDA-capable machine:
  `method = "cumcuda"` moves a coffee-break workload to real time.
  When you only need the peaks (not the grids themselves), call
  `wccCalcPeakBatch` instead of the two-step pipeline, and add
  `precision = "single"` on consumer cards.

## Session info

```{r session}
sessionInfo()
```
